Estimating the distribution of block sizes as a function of average transaction rates

### 2015-10-22 paper/paper.tex

The first part of the model is determined for us since we know $\delta_k\stackrel{iid}{\sim} Exp(1/10)$ by the design of Bitcoin, but determining the how the amount of transaction data is distributed requires more work. Crucial constraints on the models we can fit come from the data obtained from the block chain. Since Bitcoin is a distributed network without a home node, there is no official timestamp for when each block arrives on the chain. Instead, the block miner's local time is recorded. Two constraints on the reported time that the block was mined are built into the Bitcoin protocol. First, any block submitted to the chain with a local time less than the median of local times of the last 11 blocks is rejected. Second, any block with a local time greater than network adjusted time plus two hours is rejected, where network adjusted time is defined as the local time plus the median difference between local time and the time at all nodes connected to the local node with a maximum offset of 70 minutes. Because of the nature of the blocks' timestamps, occasionally a block has a timestamp which is earlier than one or more previous blocks in the chain. Essentially, we have measurement error on the measurement times, and the lower and upper bounds for valid timestamps define lower and upper bounds for the measurement error distribution. Unfortunately network adjusted time is not recorded in the blockchain, so the upper bound is not observed, but the lower bound is at our disposal.

@@

The gamma process is particularly convenient since the distribtion of $x_{\tau+\delta} - x_\tau$ and $x_\tau$ are both known in closed form as a consequence of the fact that if $x_1\sim G(\alpha_1,\beta)$ independent of $x_2\sim G(\alpha_2,\beta)$, then $x_1+x_2 \sim G(\alpha_1 + \alpha_2, \beta)$. More generally the gamma distribution is infinitely divisible so that if $x\sim G(\alpha,\beta)$ then for any integer $n>0$ there exist iid random variables $y_1,y_2,\dots,y_n$ such that $y_1+y_2+\dots+y_n$ and $x$ have the same distribution. Often L\'{e}vy processes are defined by considering an infinitely divisible distribution for $x_1$, but the distribution of $x_{\tau}$ for any $\tau>0$ need not be from the same class as $x_1$. The inverse Gaussian process is another L\'{e}vy process for which the distribution of $x_\tau$ is in the same class as the distribution for $x_1$ for all $\tau>0$. The inverse Gaussian density is
\[
p(x) = \left[\frac{\lambda}{2\pi x^3}\right]^{1/2}\exp\left[\frac{-\lambda(x - \mu)^2}{2\mu^2x}\right]
\]
for $x>0$ with mean parameter $\mu>0$, shape parameter $\lambda>0$, and variance $\mu^3/\lambda$. Then increments of the inverse Gaussian process are inverse Gaussian distributed, viz. $x_{\tau + \delta} - x_\tau \sim IG(\mu\delta,\lambda\delta^2)$.

Since the observation times are observed with error, it is attractive that we can write down the distribution of $x_{\tau + \delta} - x_\tau$ in closed form for any $tau$ and $\delta>0$. This simplifies fitting the model because it allows us to write down a relatively simple exact likelihood function, facilitating both maximum likelihood estimation and Markov chain Monte Carlo methods. The parameters of the Gamma or inverse Gaussian distributions can also be further modeled as stochastic processes in order to capture e.g. seasonal variation or stochastic volatility, though most reasonable models on these parameters will not allow for closed form distributions of the parameter increments. We will use the gamma process since it is widely available in statistical packages and in particular is available in Stan \citep{stan-software:2015}, unlike the main alternative, the inverse Gaussian.

On the data level we have measurement error with a lower bound, motivating a class of models of the form 
\[
t_{k+1} |t_{1:k}, \tau_{1:K} \stackrel{ind}{\sim} p(t|\tau_{k+1}, \phi)\mathbbm{1}\left\{t > m_{k+1})\right\}
\]
where $p(t|\tau,\phi)$ is a class of probability densities on $t>0$ with centrality parameter $\tau$ and dispersion parameter $\phi$, and $m_k = \mathrm{median}(t_{k - 0:10})$. For example $\tau$ could be the mean or the mode and $\phi$ the variance or precision. This interpretation of these parameter applies for the {\it untruncated} distribution $p(t|\tau,\phi)$, but not necessarily for the condtional dsitribution of $t_k$. A natural choice here is another gamma distribution, since the untruncated distribution still must be non-negative. However the gamma causes numerical problems at times since it forces the latent measurement times, $\tau_k$ to enter the gamma function. A similar distribution that avoids numerical problems is the lognormal, which we will use.

@@

This approach has two major shortfalls: 1) it ignores parameter uncertainty, and 2) it assumes that the other parameters will remain constant as $\gamma$ increases. In particular, 2) implies that the variance of transaction rates is always proportional to the mean, i.e. it is $\gamma/\beta$. We can mitigate 1) while making 2) if not worse, different. Specifically, given access to an approximate simulation of the posterior of the model parameters, we can simulate the model for each draw from the posterior in order to estimate the probability of a block filling up --- this is the posterior predictive distribution \citep{gelman2014bayesian}. In order to allow for different transaction rates in this simulation, we can set $\gamma$ at a particular value while using the full posterior of the other model parameters, or even better we can shift the posterior of $\gamma$ so that the posterior mean of the shifted posterior is the desired value. This approach takes into account parameter uncertainty but assumes that our uncertainty about the other parameters would be unchanged if the location of $\gamma$'s distribution were different. In other words, it still assumes that the variance of transaction rates is still $\gamma/\beta$, but now $\gamma$ and $\beta$ are also stochastic with $\beta$ distributed according to its marginal posterior distribution and $\gamma|\beta$ distributed according to a shifted version of its conditional posterior given $\beta$.

 Essentially this requires that if we were in a higher transaction rate regime, our priors would have been different such that the posterior distribution would have obtained would be the shifted posterior we construct. Explicitly constructing this prior would be involved and maybe intractable or even impossible. Instead we estimate the probability of a block filling using all three approaches, but acknowledge that the estimates become less credible for values of $\gamma$ or $\mathrm{E}[\gamma]$ further away from their estimated values.

@@

Under a fixed mean transaction rate, the probability of a full block increases as $\beta$ decreases, especially for smaller mean transaction rates. For larger mean transaction rates there is little effect on the probability the next block is full from decreasing $\beta$. This is expected since increasing the spread of transaction rates while holding the mean constant increases the variance of block sizes. When block sizes are on average much smaller than 1 MB, increasing the variance of block sizes puts significantly more probability mass above 1 MB, but when the mean is near but still below 1 MB it has a relatively small effect. What is unexpected is that taking into account the uncertainty in $\beta$ and then in both $\gamma$ and $\beta$ seems to have essentially no effect on the probability that the next block fills given a particular value for the expected value of $\gamma$.

These estimates may seem rather high considering blocks only recently began filling up. Recent transaction rates have been fluctuating around 0.05 for several months and even under constant transaction rates, the probability of a full block is over 0.1. With a constant transaction rate of 0.02, the probability of a full block is 0.067. This seems small, but in 100 blocks, which is approximately 17 hours, the probability of at least one block filling up is $1-(1-0.067)^{100}\approx 1$. The apparent paradox is solved by acknowledging a soft limit that Bitcoin miners impose on block sizes. While the Bitcoin protocol limits block sizes to 1 MB, Bitcoin miners have the ability to set their own block size limit lower. Miners have an incentive to keep blocks small since that reduces the chance of an orphan block --- a mined block which is not accepted by the blockchain --- so they tend to exercise this option. Further, the default setting is to limit block sizes to 750 KB.

>>>>

The first part of the model is determined for us since we know $\delta_k\stackrel{iid}{\sim} Exp(1/10)$ by the design of Bitcoin, but determining the how transaction data is distributed requires more work. Crucial constraints on the models we can fit come from the data obtained from the block chain. Since Bitcoin is a distributed network without a home node, there is no official timestamp for when each block arrives on the chain. Instead, the block miner's local time is recorded. Two constraints on the reported time that the block was mined are built into the Bitcoin protocol. First, any block submitted to the chain with a local time less than the median of local times of the last 11 blocks is rejected. Second, any block with a local time greater than network adjusted time plus two hours is rejected, where network adjusted time is defined as the local time plus the median difference between local time and the time at all nodes connected to the local node with a maximum offset of 70 minutes. Because of the nature of the blocks' timestamps, occasionally a block has a timestamp which is earlier than one or more previous blocks in the chain. Essentially, we have measurement error on the measurement times, and the lower and upper bounds for valid timestamps define lower and upper bounds for the measurement error distribution. Unfortunately network adjusted time is not recorded in the blockchain, so the upper bound is not observed, but the lower bound is at our disposal.

@@

The gamma process is particularly convenient since the distribution of $x_{\tau+\delta} - x_\tau$ and $x_\tau$ are both known in closed form as a consequence of the fact that if $x_1\sim G(\alpha_1,\beta)$ independent of $x_2\sim G(\alpha_2,\beta)$, then $x_1+x_2 \sim G(\alpha_1 + \alpha_2, \beta)$. More generally the gamma distribution is infinitely divisible so that if $x\sim G(\alpha,\beta)$ then for any integer $n>0$ there exist iid random variables $y_1,y_2,\dots,y_n$ such that $y_1+y_2+\dots+y_n$ and $x$ have the same distribution. Often L\'{e}vy processes are defined by considering an infinitely divisible distribution for $x_1$, but the distribution of $x_{\tau}$ for any $\tau>0$ need not be from the same class as $x_1$. The inverse Gaussian process is another L\'{e}vy process for which the distribution of $x_\tau$ is in the same class as the distribution for $x_1$ for all $\tau>0$. The inverse Gaussian density is
\[
p(x) = \left[\frac{\lambda}{2\pi x^3}\right]^{1/2}\exp\left[\frac{-\lambda(x - \mu)^2}{2\mu^2x}\right]
\]
for $x>0$ with mean parameter $\mu>0$, shape parameter $\lambda>0$, and variance $\mu^3/\lambda$. Then increments of the inverse Gaussian process are inverse Gaussian distributed, viz. $x_{\tau + \delta} - x_\tau \sim IG(\mu\delta,\lambda\delta^2)$.

Since the observation times are observed with error, it is attractive that we can write down the distribution of $x_{\tau + \delta} - x_\tau$ in closed form for any $tau$ and $\delta>0$. This simplifies fitting the model because it allows us to write down a relatively simple exact likelihood function, facilitating both maximum likelihood estimation and Markov chain Monte Carlo methods. The parameters of the Gamma or inverse Gaussian distributions can also be further modeled as stochastic processes in order to capture e.g. seasonal variation or stochastic volatility, though most reasonable models on these parameters will not allow for closed form distributions of the transaction data increments. We will use the gamma process since it is widely available in statistical packages and in particular is available in Stan \citep{stan-software:2015}, unlike the main alternative, the inverse Gaussian.

On the data level we have measurement error with a lower bound, motivating a class of models of the form 
\[
t_{k+1} |t_{1:k}, \tau_{1:K} \stackrel{ind}{\sim} p(t|\tau_{k+1}, \phi)\mathbbm{1}\left\{t > m_{k+1})\right\}
\]
where $p(t|\tau,\phi)$ is a class of probability densities on $t>0$ with centrality parameter $\tau$ and dispersion parameter $\phi$, and $m_k = \mathrm{median}(t_{k - 0:10})$. For example $\tau$ could be the mean or the mode and $\phi$ the variance or precision. This interpretation of these parameter applies for the {\it untruncated} distribution $p(t|\tau,\phi)$, but not necessarily for the conditional dsitribution of $t_k$. A natural choice here is another gamma distribution, since the untruncated distribution still must be non-negative. However the gamma causes numerical problems at times since it forces the latent measurement times, $\tau_k$ to enter the gamma function. A similar distribution that avoids numerical problems is the lognormal, which we will use.

@@

This approach has two major shortfalls: 1) it ignores parameter uncertainty, and 2) it assumes that the other parameters will remain constant as $\gamma$ increases. In particular, 2) implies that the variance of transaction rates is always proportional to the mean, i.e. it is $\gamma/\beta$. We can mitigate 1) while making 2) if not worse, different. Specifically, given access to an approximate simulation of the posterior of the model parameters, we can simulate the model for each draw from the posterior in order to estimate the probability of a block filling up --- this is the posterior predictive distribution \citep{gelman2014bayesian}. In order to allow for different transaction rates in this simulation, we can set $\gamma$ at a particular value while using the full posterior of the other model parameters, or even better we can shift the posterior of $\gamma$ so that the posterior mean of the shifted posterior is the desired value. This approach takes into account parameter uncertainty but assumes that our uncertainty about the other parameters would be unchanged if the location of $\gamma$'s distribution were different. In other words, it still assumes that the variance of transaction rates is still $\gamma/\beta$, but now $\gamma$ and $\beta$ are also stochastic with $\beta$ distributed according to its marginal posterior distribution and $\gamma|\beta$ distributed according to a shifted version of its conditional posterior given $\beta$. Essentially this assumption requires that if we were in a higher transaction rate regime, our priors would have been different such that the posterior distribution we would have obtained would be the shifted posterior we construct. Explicitly constructing this prior would be involved and maybe intractable or even theoretically impossible. Instead we estimate the probability of a block filling using all three approaches, but acknowledge that the estimates become less credible for values of $\gamma$ or $\mathrm{E}[\gamma]$ further away from their estimated values.

@@

Under a fixed mean transaction rate, the probability of a full block increases as $\beta$ decreases, especially for smaller mean transaction rates. For larger mean transaction rates there is little effect on the probability the next block is full from decreasing $\beta$. This is expected since increasing the spread of transaction rates while holding the mean constant increases the variance of block sizes. When block sizes are on average much smaller than 1 MB, increasing the variance of block sizes puts significantly more probability mass above 1 MB, but when the mean is near but still below 1 MB it has a relatively small effect. What is unexpected is that no matter what the expected mean transaction rate is, the probability of a block filling up is largely unaffected by taking into account parameter uncertainty in $\beta$ or in both $\beta$ and $\gamma$, which can be seen from the final three lines of Table \ref{tab:probsims}.

%% In the next draft we can see what happens when the mean transaction rate varies deterministically over time -- e.g. through some prespecified seasonal pattern, then see how that impacts not the probability of the next block filling up, but one of the next 10 blocks, for example.

%% These estimates seem rather high considering no block have filled up yet. Transaction rates have been fluctuating around 0.05 for several months and even under constant transaction rates, the probability of a full block is over 0.1. Even at a constant transaction rate of 0.02, the probability of a full block is 0.067. In 100 blocks, which is approximately 17 hours, the probability of at least one block filling up is $1-(1-0.067)^{100}\approx 1$. Since no block has filled up yet, clearly something is happening that prevents blocks from filling up. One possibility is that transaction rates are usually low and in no danger of causing a full block, but occasionally there are large spikes in the rate. Another possibility is strategic action by Bitcoin traders and miners to prevent blocks from filling up. Traders may be waiting to trade until a new block replaces a nearly full block, which is not much of a cost since the wait is only about 10 minutes on average. Similarly miners have an incentive to keep blocks small since that reduces the change of an orphan block --- a mined block which is not accepted by the blockchain --- so the miner of the current block may be rejecting some transactions in order to prevent the limit from being reached.

%% [A PLACEHOLDER FOR NEXT DRAFT IN WHICH WE DISCUSS SOFT LIMITS AND HOW THEY AFFECT OUR MODEL?]   Matt: sounds good.

vvvvvvvvvvvvvvvvvvvv
\section{Discussion}
Our analysis does not resolve the contentious block size issue, but it is an important input into weighing the tradeoff between supporting higher transaction rates and the security issues created by larger blocks. For example, at the existing 1 MB maximum, as average transaction rates increase from 500 KB/block to 600 KB/block, an additional five percent or so of blocks will hit the limit. These full blocks signal that not all transactions are being included in the first block after their transmission.

In addition, a greater understanding of the dynamics of full blocks versus the average transaction rates could be used to help clients set fees appropriately, as a function of the observed rate of transactions over the last several blocks and the existing maximum block size. Creating a more dynamic fee market is another way, besides simply increasing the maximum block size, to scale the network effectively.
^^^^^^^^^^^^^^^^^^^^
